---
title: "Introduction to R"
format: html
editor: visual
---

## Overview

This is a Quarto document - you may see people refer to this as a notebook. This type of file allows you to weave text, code, tables, and charts together.

You can output a variety of file types including html and pdf for sharing with others. It has the file extension `.qmd`. This contrasts to the `.R` script file we saw previously which only contains code.

To style the document we can use markdown, a lightweight markup language designed for quick, easy formatting. You can find more about markdown [here](https://quarto.org/docs/authoring/markdown-basics.html).

The rest of this course will use this notebook.

## Libraries

```{r}

```

## Reading data with readr

Constructing the file names.

```{r}

```

A good check to do is to see if the file you have created exists.

```{r}

```

Now we can read it the data files

```{r}

```

Let's try some of those functions that the output suggests.

```{r}

```

```{r}

```

We can use the viewer or the print function.

```{r}

```

Using the summary function.

```{r}

```

Turn off scientific notation.

```{r}

```

Read in the population data

```{r}

```

## Manipulating data with dplyr

Manipulate the GDP data

```{r}

```

Manipulate the population and combine.

```{r}
population_edited <- population_raw |> 
  arrange(year) |> 
  select(population)

combined <- bind_cols(gdp_edited, population_edited) |> 
  mutate(gdp_per_head = annual_gdp / population)
```

## Plotting data with ggplot2

```{r}

```

## Output

How to save those charts outside of R?

```{r}

```

Saving the data outside of R.

```{r}

```
